
---
title: Statistical analysis of the count matrix of extracted reactions
author: Jan Taubenheim
date: "`r Sys.Date()`"
output: 
        html_document:
                toc: true
                theme: united
---

```{r loadLibraries}

knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.width = 10, fig.height=5)

require(ggplot2)
require(data.table)
require(lme4)
require(lmerTest)
require(lqmm)
require(car)
require(cowplot)
require(DT)
require(pheatmap)
require(limma)
require(cluster)

```

```{r debug, eval=TRUE, include =FALSE}

# This section is only for debugging purposes
setClass("snakemake", representation(input = "list", output = "list"))
snakemake <- new("snakemake", 
                    input = list(
                        rxnCount = "../results/data/ModelAnalysis/MatjesAbsorption.colormore22-rxnIdMat.csv",
                        subSysCount = "../results/data/ModelAnalysis/MatjesAbsorption.colormore22-SubSysMat.csv",
                        lmmModels = "../results/FVA/LMMmodelsRangeFVA_MatjesAbsorption.colormore22-*Diagnosis_zeroBiomass.csv",
                        rxn_stats = "../results/FVA/PERMANOVAmodels_MatjesAbsorption.colormore22-distFVA_zeroBiomassvs~HB_Mayo_impu*Diagnosis.csv",
                        fva = "../results/FVA/FVA_MatjesAbsorption.colormore22-zeroBiomass.RDS",
                        distFiles = "../results/FVA/Rxndist_MatjesAbsorption.colormore22_zeroBiomass.tar.gz",
                        meta = "../resources/META_emed_future.csv",
                        clinic = "../resources/ClinicalData.csv",
                        subsysTable = "../resources/Subsystems.csv",
                        sampleBlacklist = "../resources/sampleBlacklist.csv",
                        GPR = "../resources/colormore22_GPR.csv"
                    ),
                    output = list(
                        html = "results/reports/MatjesAbsorption.colormore22-Natrix_analyseRxnCounts.html"
                    )
                )

```

```{r loadData}

meta <- fread(snakemake@input[["meta"]])
meta[,PatientID_Time := paste(PatientID, Time_seq, sep ="_")]
meta <- meta[Time_seq != -1,]
clinic <- fread(snakemake@input[["clinic"]], drop=1)
meta <- merge(meta,clinic, by="PatientID_Time", all.x=TRUE, suffix = c(".meta",""))
#meta_merge <- merge(meta,clinic, all.x=TRUE)

subSys <- fread(snakemake@input[["subSysCount"]])
subSysCount <- as.matrix(subSys[,-1, with =F])
rownames(subSysCount) <- gsub(" ","_",unlist(subSys[,1]))

rxns <- fread(snakemake@input[["rxnCount"]])
rxnCount <- as.matrix(rxns[,-1,with=F ])
rownames(rxnCount) <- gsub(" ","_",unlist(rxns[,1]))

subSystems <- fread(snakemake@input[["subsysTable"]])
GPR <- fread(snakemake@input[["GPR"]])

# remove degraded samples
blcklst <- fread(snakemake@input[["sampleBlacklist"]])
blckSmpls <- blcklst[relevance < 2, Sample]
meta <- meta[!(SeqID %in% blckSmpls),]
subSysCount <- subSysCount[,!(colnames(rxnCount) %in% blckSmpls)]
rxnCount <- rxnCount[,!(colnames(rxnCount) %in% blckSmpls)]
meta <- meta[SeqID %in% colnames(rxnCount),]

lmmModels <- fread(snakemake@input[["lmmModels"]])
rxnDistStats <- fread(snakemake@input[["rxn_stats"]])
fva <- readRDS(snakemake@input[["fva"]])
distTar <- snakemake@input[["distFiles"]]

```


# Find Diagnosis specific reactions

The idea is to find reactions which are specifically associated with the HB/Mayo score within on diagnosis (UC or CD).

```{r venn}

vennMat <- matrix(FALSE, nrow = length(unique(rxnDistStats[,Rxn])), ncol = 2, dimnames = list(unique(rxnDistStats[,Rxn]),c("PERMANOVA","LMM")))
vennMat[lmmModels[padj <= 0.05 & variable %in% grep(":Diagnosis", variable, value = TRUE), gsub(":Diagnosis","",variable)],"LMM"] <- TRUE
vennMat[rxnDistStats[padj <= 0.05 & Variable %in% grep(":Diagnosis", Variable, value = TRUE), Rxn],"PERMANOVA"] <- TRUE
vennDiagram(vennMat, main = "Diagnosis specific reactions associated with HB/Mayo")

```

There is quite some discrepancy between the two different methods - this is surprising, but okay...

```{r plots}

rxns <- rownames(vennMat)[apply(vennMat,1,all)]
fva_plot <- data.table(reshape2::melt(fva[rxns,,]))
fva_plot <- merge(fva_plot,meta[Tissue=="Biopsy",],by.x="sample",by.y="SeqID")

ggplot(fva_plot[Diagnosis == "CU",], aes(x=HB_Mayo_impu, y=value, color = Var3)) +
    facet_wrap(~rxn, scale = "free") +
    geom_point()+
    geom_smooth() +
    theme_bw() +
    ggtitle("Diagnosis:UC")

ggplot(fva_plot[Diagnosis == "CD",], aes(x=HB_Mayo_impu, y=value, color = Var3)) +
    facet_wrap(~rxn, scale = "free") +
    geom_point()+
    geom_smooth()+
    theme_bw() +
    ggtitle("Diagnosis:CD")

```

These reactions are all associated to the fatty acid exchange in short: Carnitine import is reduced in higher inflammation in CU, additionally import and conversion of acyl-carnitine to acyl-CoA within the ER and peroxisomes are reduced. Whatever this means for the treatment. Additionally there is an increased conversion from dihydrofolate to tetrahydrofolate at higher inflammation - this reaction is a target for methotrexate (used for immunsupression).




