---
title: Reaction expression score analysis for IBD patients 
author: Jan Taubenheim
date: "`r Sys.Date()`"
output: 
    rmdformats::downcute:
        highlight: tango
        use_bookdown: TRUE
bibliography: ["/home/taube/Documents/References/Bibtex/My_Library.bib"]
link-citations: true
---


```{r knitrSetup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.width = 10, fig.height=5)
```

```{r loadLibraries}

require(data.table)
require(lme4)
require(lmerTest)
require(ggplot2)
require(cowplot)
require(clusterProfiler)
require(ComplexUpset)

```

```{r loadData}

subsystems <- fread("./Statistics/dat/subsystems.csv")
cluster <- fread("./Statistics/results/rxnExpr_DBSCAN_Cluster.csv")
stats <- fread("./Statistics/results/rxnExpr_coefficients_fullTable.csv")
stats.padj <- stats[Coef != "(Intercept)",]
stats.padj[,padj := p.adjust(`Pr(>|t|)`, method = "BH")]
sigs <- stats.padj[padj < 0.05,]

```

# Introduction

I created context specific models of different IBD patients (longitudinal data) from biopsy samples. Afterwards I used FVA results to gain insights into the reactions which do significantly change with differing disease activities expressed as HB/Mayo scores. However, the first step in the pipeline was the mapping of expression values from genes to the reactions in the network. This can be also used for analysis, which I will do here. To this end I used linear mixed models to explain the HB/Mayo score by the change of reaction expression scores for each reaction.

# Results

Again I used a clustering to de-correlate the reactions and to reduce the amount of tests. 


## Clustering 

It might be a good idea to understand what the clustering has done to the data

```{r clusteringPlots}

cluster.size <- cluster[cluster != 0, .(cluster_size = .N), by=rep.rxn]
cluster.size.hist <- ggplot(cluster.size, aes(x=cluster_size)) +
    geom_histogram(binwidth = 10) +
    labs(x= "Cluster size", y= "Counts", title = "Number of reactions in cluster")+
    theme_bw()

cluster.size.ecdf <- ggplot(cluster.size, aes(x=cluster_size)) +
    stat_ecdf()+
    labs(x= "Cluster size", y= "Fraction of data covered", title = "Number of reactions in cluster")+
    theme_bw()

cowplot::plot_grid(cluster.size.hist,
                   cluster.size.ecdf,
                   ncol = 2)

```


There are a lot of small clusters with only a few members, while there are also larger ones. But generally, that means clustering was done only on highly correlated reactions. A total of `r nrow(cluster.size)` has been generated.

```{r clusterSmry}


clusterSummary <- data.frame(
                             rxns = nrow(cluster),
                             cluster = max(cluster[,cluster]),
                             rxns_in_cluster = nrow(cluster[cluster != 0,]),
                             reduction = round((max(cluster[,cluster])+nrow(cluster[cluster == 0, ]))/ nrow(cluster),2),
                             significant_rxns_and_cluster = length(unique((sigs[,rxn]))),
                             significant_rxns = nrow(cluster[rep.rxn %in% sigs[,rxn],]), #
                             largest_cluster = max(cluster.size[,cluster_size]),
                             smallest_cluster = min(cluster.size[,cluster_size]),
                             mean_cluster = round(mean(cluster.size[,cluster_size]),2),
                             median_cluster = median(cluster.size[,cluster_size])
                             )
clusterSummary <- t(clusterSummary)
colnames(clusterSummary) <- c("Value")
DT::datatable(clusterSummary)

```

There are a lot of clusters which are associated to the HB/Mayo change.

## Association with HB/Mayo

### Overview

```{r histPvals}

sigs.hist <- ggplot(sigs, aes(x=Estimate)) + 
    geom_histogram(binwidth=0.1)+
    labs(x = "Coefficient for change in HB/Mayo if range/center is increased",
         y = "Counts",
         fill = "Change in") + 
    theme_bw()

sigs.hist

```

Similar to the presence/absence of reactions, the histogram is shifted to the left. Again this speaks to a less flexible metabolism in higher inflammatory states.

### Subsystems association

To get an overview of what subsystems are associated with the change in HB/Mayo we can associate the different subsystems to the significant reactions.

```{r subsystems}

# get all significant reactions
rxns <- unique(cluster[rep.rxn %in% sigs[,rxn], rxn])


# get the annotation for these reactions
setkey(subsystems, "reaction")
subs <- subsystems[rxns]

#merge the stats with the reactions - use the clusterID to merge as well
subs <- merge(subs, cluster, by.x = "reaction", by.y = "rxn", all.x=T)
subs <- merge(subs, sigs, by.x = "rep.rxn", by.y = "rxn") 

# create a boxplot of the coefficients per subsystem
subs_order <- subs[,.(mean_est = mean(Estimate)), by = subsystem]
subs[,subsystems := factor(subsystem, levels=unique(subs_order[order(mean_est,decreasing=TRUE), subsystem]))]

subs_jitterplot <- ggplot(subs, aes(x=subsystems, y= Estimate)) +
        geom_hline(yintercept=0, color = "red", linetype=2)+
        geom_boxplot(outlier.shape = NA)+
        geom_jitter(height=0, width=0.3)+
        theme_bw()+
        theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
subs_jitterplot

```

### Subsystem enrichment

In order to test if and which of the subsystems are significantly associated I will perform an hypergeometric enrichment analysis. Note that I will use simply all reactions which are either significantly associated with the center or the range or both to HB/Mayo values.

```{r geomEnrichment, fig.height = 8}

hgt <- enricher(unique(sigs[,rxn]), TERM2GENE = subsystems)
dotplot(hgt,x="Count", showCategory = nrow(data.frame(hgt))) +
        labs(x="Number of reactions", size = "RxnRatio") 


```

Additionally to the hypergeometric enrichment I will perform a gene set enrichment analysis (GSEA).

```{r GSEA}

gsea <- readRDS("Statistics/results/rxnExpr_GSEASubsystems.RDS")
ridgeplot(gsea, label_format=90) + theme(axis.text.y= element_text(size = 12))

```

### Comparing approachResponsees

Again there is overlap in the methods, so we can have a look at this.

```{r upsetplot}

# get the subsystems of each analysis
subsystems.names <- unique(subsystems[,subsystem])
tbl.gsea <- data.table(as.data.frame(gsea))
tbl.hgt <- data.table(as.data.frame(hgt))

mat <- matrix(FALSE, nrow = length(subsystems.names), ncol = 3, dimnames = list(x=subsystems.names, y=c("Counts","HGT","GSEA")))

mat[unique(subs[!is.na(subsystem),subsystem]),"Counts"] <- TRUE
mat[tbl.gsea[,ID], "GSEA"] <- TRUE
mat[tbl.hgt[,ID], "HGT"] <- TRUE

upset(as.data.frame(mat), intersect = colnames(mat))

```

Interestingly, there are `r nrow(mat[which(rowSums(mat) == 1 & mat[,"GSEA"]==1),])` subsystems which contain no single reaction which can be associated to HB/Mayo on its own, but in total contribute to the change in HB/Mayo. These are the following:`r names(which(rowSums(mat) == 1 & mat[,"GSEA"]==1))`. Furthermore, there are `r length(which(rowSums(mat)==2 & rowSums(mat[, c("Counts","HGT")]) == 2))` subsystems which are only present in the HGT analysis. These are: `r names(which(rowSums(mat)==2 & rowSums(mat[, c("Counts","HGT")]) == 2))`.

In total there are `r nrow(mat[which(rowSums(mat) == ncol(mat)),])` subsystems which can be found as overlap in all analysis: `r rownames(mat[which(rowSums(mat) == ncol(mat)),])`.


## Association with Response

A second round of testing included those reactions which are associated to the response of the patients to treatment of IBD. The association was corrected with the HB/Mayo score, to obtain reactions which show a good correlation to the response even if they not necessarily correlate with the HB/Mayo score.

```{r loadDataResponse}

cluster <- fread("./Statistics/results/rxnExpr_DBSCAN_Cluster.csv")
stats.padj <- fread("./Statistics/results/rxnExpr_GLMMsResponseAll_coefs.csv")
sigs <- stats.padj[padj < 0.05,]

```


### Overview

```{r histPvalsResponse}

sigs.hist <- ggplot(sigs, aes(x=Estimate)) + 
    geom_histogram(binwidth=0.1)+
    labs(x = "Coefficient for change in HB/Mayo if range/center is increased",
         y = "Counts",
         fill = "Change in") + 
    theme_bw()
sigs.hist

```

The histogram looks pretty symmetric.

### Subsystems association

To get an overview of what subsystems are associated with the change in Response we can associate the different subsystems to the significant reactions.

```{r subsystemsResponse}

# get all significant reactions
rxns <- unique(cluster[rep.rxn %in% sigs[,rxn], rxn])


# get the annotation for these reactions
setkey(subsystems, "reaction")
subs <- subsystems[rxns]

#merge the stats with the reactions - use the clusterID to merge as well
subs <- merge(subs, cluster, by.x = "reaction", by.y = "rxn", all.x=T)
subs <- merge(subs, sigs, by.x = "rep.rxn", by.y = "rxn") 

# create a boxplot of the coefficients per subsystem
subs_order <- subs[,.(mean_est = mean(Estimate)), by = subsystem]
subs[,subsystems := factor(subsystem, levels=unique(subs_order[order(mean_est,decreasing=TRUE), subsystem]))]

subs_jitterplot <- ggplot(subs, aes(x=subsystems, y= Estimate)) +
        geom_hline(yintercept=0, color = "red", linetype=2)+
        geom_boxplot(outlier.shape = NA)+
        geom_jitter(height=0, width=0.3)+
        theme_bw()+
        theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
subs_jitterplot

```

### Subsystem enrichment

In order to test if and which of the subsystems are significantly associated I will perform an hypergeometric enrichment analysis. Note that I will use simply all reactions which are either significantly associated with the center or the range or both to the Response to treatment in these patients.

```{r geomEnrichmentResponse, fig.height = 8}

hgt <- enricher(unique(sigs[,rxn]), TERM2GENE = subsystems)
dotplot(hgt,x="Count", showCategory = nrow(data.frame(hgt))) +
        labs(x="Number of reactions", size = "RxnRatio") 


```

Additionally to the hypergeometric enrichment I will perform a gene set enrichment analysis (GSEA).

```{r GSEAResponse}

gsea <- readRDS("Statistics/results/rxnExpr_GSEASubsystems_Response.RDS")
ridgeplot(gsea, label_format=90) + theme(axis.text.y= element_text(size = 12))

```

### Comparing approaches

Again there is overlap in the methods, so we can have a look at this.

```{r upsetplotResponse}

# get the subsystems of each analysis
subsystems.names <- unique(subsystems[,subsystem])
tbl.gsea <- data.table(as.data.frame(gsea))
tbl.hgt <- data.table(as.data.frame(hgt))

mat <- matrix(FALSE, nrow = length(subsystems.names), ncol = 3, dimnames = list(x=subsystems.names, y=c("Counts","HGT","GSEA")))

mat[unique(subs[!is.na(subsystem),subsystem]),"Counts"] <- TRUE
mat[tbl.gsea[,ID], "GSEA"] <- TRUE
mat[tbl.hgt[,ID], "HGT"] <- TRUE

upset(as.data.frame(mat), intersect = colnames(mat))

```

Interestingly, there are `r nrow(mat[which(rowSums(mat) == 1 & mat[,"GSEA"]==1),])` subsystems which contain no single reaction which can be associated to Response on its own, but in total contribute to the change in Response. These are the following:`r names(which(rowSums(mat) == 1 & mat[,"GSEA"]==1))`. Furthermore, there are `r length(which(rowSums(mat)==2 & rowSums(mat[, c("Counts","HGT")]) == 2))` subsystems which are only present in the HGT analysis. These are: `r names(which(rowSums(mat)==2 & rowSums(mat[, c("Counts","HGT")]) == 2))`.

In total there are `r nrow(mat[which(rowSums(mat) == ncol(mat)),])` subsystems which can be found as overlap in all analysis: `r rownames(mat[which(rowSums(mat) == ncol(mat)),])`.
