---
title: Analysis of day 14
author: Jan Taubenheim
date: "`r Sys.Date()`"
output: 
        html_document:
                toc: true
                theme: united
---


```{r knitrSetup}

knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.width = 10, fig.height=5)
```

```{r libraries}

require(ggplot2)
require(data.table)
require(FactoMineR)
require(factoextra)
require(nnet)
require(car)

```


```{r loadData}

fva <- readRDS("../results/data/FVA/MatjesAbsorption.colormore22-FVA_zeroBiomass.RDS")
fvaDist <- "../results/data/FVA/MatjesAbsorption.colormore22_zeroBiomass_Rxndist.tar.gz"
fba <- read.csv("../results/data/FBA/MatjesAbsorption.colormore22-fluxes.csv", row.names=1)

meta <- fread("../resources/META_emed_future.csv")
meta[,PatientID_Time := paste(PatientID, Time_seq, sep ="_")]
meta <- meta[Time_seq != -1,]
clinic <- fread("../resources/ClinicalData.csv", drop=1)
meta <- merge(meta,clinic, by="PatientID_Time", all.x=TRUE, suffix = c(".meta",""))


# remove degraded samples
blcklst <- fread("../resources/sampleBlacklist.csv")
blckSmpls <- blcklst[relevance < 2, Sample]
#blckSmpls <- c(blckSmpls, "H18597.L1_S29_L003","H18606.L1_S38_L004")
meta <- meta[!(SeqID %in% blckSmpls),]

```


# Primer

To get things clear here a plot how the HB/Mayo behaves over time.

```{r HB_Mayo}


pp <- ggplot(meta, aes(x=Time_seq, y = HB_Mayo_impu, color = Response)) +
    geom_point() +
    geom_smooth(method="glm",method.args = list(family="quasipoisson")) +
    facet_wrap(~Diagnosis)
pp

```


# PCA

Lets do the PCA analysis only on the 14days data for the FVA range.

```{r PCA14d}

pca_dat <- fva[,meta[Time_seq == 14 & Tissue == "Biopsy", SeqID], "range"]
pca_dat <- pca_dat[apply(pca_dat,1,var) > 1E-6,]
# remove highly correlated reactions
pca_cor <- cor(t(pca_dat))
pca_cor <- is.na(pca_cor) | abs(pca_cor) > 0.8
pca_cor <- cbind(as.data.frame(t(combn(rownames(pca_cor),2))), cor = pca_cor[lower.tri(pca_cor)])
pca_cor <- data.table(pca_cor)
sel <- pca_cor[,any(cor),by=V2]
sel <-sel[V1 == FALSE,V2]

pca_dat <- pca_dat[sel,]

pca <- prcomp(t(pca_dat),scale=TRUE, center=TRUE)
pca_d <- merge(data.table(pca[["x"]], keep.rownames=TRUE), meta, by.x="rn",by.y="SeqID")

contrib  <- sapply(1:ncol(pca_dat), function(x) {
                       pca[["rotation"]][,x] * (pca_dat[,x] - summary(pca)[["center"]])/summary(pca)[["scale"]]
})
colnames(contrib) <- paste0("PC", 1:ncol(contrib))

contribOrder <- apply(contrib, 2, function(x) rownames(contrib)[order(abs(x), decreasing=TRUE)])


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = HB_Mayo_impu <=5)) +
    geom_point(size=3, alpha=0.5) +
    facet_wrap(~Diagnosis)
pp


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = Response)) +
    geom_point(size=3, alpha=0.5) + 
    facet_wrap(~Diagnosis) 
pp

setkey(meta, "SeqID")
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), HB_Mayo_impu <= 5],
                geom.ind = c("point"))
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), Response],
                geom.ind = c("point"))


```

Not so much to see here. There is a slight clustering visible in terms of controls sticking closer together, but a clear distinction is missing.

Lets look across all time points once more to get an idea about the data set.


```{r PCAtimeSplit}

pca_dat <- fva[,meta[SeqID %in% colnames(fva) & Tissue == "Biopsy", SeqID], "range"]
pca_dat <- pca_dat[apply(pca_dat,1,var) > 1E-6,]

# remove highly correlated reactions
pca_cor <- cor(t(pca_dat))
pca_cor <- is.na(pca_cor) | abs(pca_cor) > 0.8
pca_cor <- cbind(as.data.frame(t(combn(rownames(pca_cor),2))), cor = pca_cor[lower.tri(pca_cor)])
pca_cor <- data.table(pca_cor)
sel <- pca_cor[,any(cor),by=V2]
sel <-sel[V1 == FALSE,V2]

pca_dat <- pca_dat[sel,]


pca <- prcomp(t(pca_dat),scale=TRUE, center=TRUE)
pca_d <- merge(data.table(pca[["x"]], keep.rownames=TRUE), meta, by.x="rn",by.y="SeqID")


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = HB_Mayo_impu <=5)) +
    geom_point(size=3, alpha=0.5) +
    facet_wrap(~Diagnosis,scale="free")
pp



pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = Response)) +
    geom_point(size=3, alpha=0.5) + 
    facet_wrap(~Diagnosis, scale="free") 
pp


setkey(meta, "SeqID")
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), HB_Mayo_impu <= 5],
                geom.ind = c("point"))
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), Response],
                geom.ind = c("point"))

pca_d <- data.table(reshape2::melt(pca[["x"]][,1:10], varnames=c("SeqID","PC")))
pca_d <- merge(pca_d,meta, by = "SeqID")
pp <- ggplot(pca_d, aes(y=value, x = HB_Mayo_impu)) +
    geom_point() +
    geom_smooth(method="loess") +
    facet_wrap(PC~Diagnosis)
pp

pp <- ggplot(pca_d[PC =="PC1",], aes(x=value, y = HB_Mayo_impu)) +
    geom_point() +
    geom_smooth(method="glm", method.args = list(family = poisson )) +
    facet_wrap(~Diagnosis, scale = "free")
pp

pp <- ggplot(pca_d[PC =="PC1",], aes(x=value, y = HB_Mayo_impu, color = Response)) +
    geom_point() +
    geom_smooth(method="glm", method.args = list(family = poisson )) +
    facet_wrap(~Diagnosis, scale = "free")
pp

pp + facet_wrap(~PatientID, scale = "free")




```

OK this is interesting and there seems to be a rather CU specific signal in the data. I will check on this specifically.

```{r PCA_CUonly}

pca_dat <- fva[,meta[SeqID %in% colnames(fva) & Tissue == "Biopsy" & Diagnosis == "CU", SeqID], "range"]
pca_dat <- pca_dat[apply(pca_dat,1,var) > 1E-6,]

# remove highly correlated reactions
pca_cor <- cor(t(pca_dat))
pca_cor <- is.na(pca_cor) | abs(pca_cor) > 0.8
pca_cor <- cbind(as.data.frame(t(combn(rownames(pca_cor),2))), cor = pca_cor[lower.tri(pca_cor)])
pca_cor <- data.table(pca_cor)
sel <- pca_cor[,any(cor),by=V2]
sel <-sel[V1 == FALSE,V2]

pca_dat <- pca_dat[sel,]


pca <- prcomp(t(pca_dat),scale=TRUE, center=TRUE)
pca_d <- merge(data.table(pca[["x"]], keep.rownames=TRUE), meta, by.x="rn",by.y="SeqID")


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = HB_Mayo_impu <=5)) +
    geom_point(size=3, alpha=0.5) 
pp
pp +
    facet_wrap(~factor(Time_seq), scale="free")


pp <- ggplot(pca_d, aes(x=PC1,y=PC5, color = Response)) +
    geom_point(size=3, alpha=0.5) 
pp
pp + 
    facet_wrap(~factor(Time_seq), scale="free")


setkey(meta, "SeqID")
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), HB_Mayo_impu <= 5],
                geom.ind = c("point"))
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), Response],
                geom.ind = c("point"))


```


# PCA with min max data

One could try to not transform the data in any way and simply use the data as is.

```{r PCAMinMax}


pca_dat <- rbind(fva[,meta[SeqID %in% colnames(fva) & Tissue == "Biopsy", SeqID], "min"],fva[,meta[SeqID %in% colnames(fva) & Tissue == "Biopsy", SeqID], "max"])
rownames(pca_dat) <- paste0(rownames(pca_dat), c(rep("_min",dim(fva)[[1]]),rep("_max",dim(fva)[[1]])))
pca_dat <- pca_dat[apply(pca_dat,1,var) > 1E-6,]

# remove highly correlated reactions
pca_cor <- cor(t(pca_dat))
pca_cor <- is.na(pca_cor) | abs(pca_cor) > 0.8
pca_cor <- cbind(as.data.frame(t(combn(rownames(pca_cor),2))), cor = pca_cor[lower.tri(pca_cor)])
pca_cor <- data.table(pca_cor)
sel <- pca_cor[,any(cor),by=V2]
sel <-sel[V1 == FALSE,V2]

pca_dat <- pca_dat[sel,]

pca <- prcomp(t(pca_dat),scale=TRUE, center=TRUE)
pca_d <- merge(data.table(pca[["x"]], keep.rownames=TRUE), meta, by.x="rn",by.y="SeqID")


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = HB_Mayo_impu <=5)) +
    geom_point(size=3, alpha=0.5) +
    facet_wrap(~Diagnosis,scale="free")
pp
pp + facet_grid(Diagnosis~factor(Time_seq))


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = Response)) +
    geom_point(size=3, alpha=0.5) + 
    facet_wrap(~Diagnosis, scale="free") 
pp
pp + facet_grid(Diagnosis~factor(Time_seq))

setkey(meta, "SeqID")
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), HB_Mayo_impu <= 5],
                geom.ind = c("point"))
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), Response],
                geom.ind = c("point"))

pca_d <- data.table(reshape2::melt(pca[["x"]][,1:10], varnames=c("SeqID","PC")))
pca_d <- merge(pca_d,meta, by = "SeqID")

pp <- ggplot(pca_d, aes(y=value, x = HB_Mayo_impu)) +
    geom_point() +
    geom_smooth(method="loess", k = 10) +
    facet_wrap(PC~Diagnosis)
pp

```


Ok there seems to be a little more information in that kind of data. 

# PCA FBA

```{r PCA_FBA}

pca_dat <- fba[,meta[SeqID %in% colnames(fba) & Tissue == "Biopsy",SeqID]]
pca_dat <- pca_dat[apply(pca_dat,1,var) > 1E-6,]

# remove highly correlated reactions
pca_cor <- cor(t(pca_dat))
pca_cor <- is.na(pca_cor) | abs(pca_cor) > 0.8
pca_cor <- cbind(as.data.frame(t(combn(rownames(pca_cor),2))), cor = pca_cor[lower.tri(pca_cor)])
pca_cor <- data.table(pca_cor)
sel <- pca_cor[,any(cor),by=V2]
sel <-sel[V1 == FALSE,V2]

pca_dat <- pca_dat[sel,]

pca <- prcomp(t(pca_dat),scale=TRUE, center=TRUE)
fviz_screeplot(pca)
pca_d <- merge(data.table(pca[["x"]], keep.rownames=TRUE), meta, by.x="rn",by.y="SeqID")


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = HB_Mayo_impu <=5)) +
    geom_point(size=3, alpha=0.5) +
    facet_wrap(~Diagnosis,scale="free")
pp
pp + facet_grid(Diagnosis~factor(Time_seq))


pp <- ggplot(pca_d, aes(x=PC1,y=PC2, color = Response)) +
    geom_point(size=3, alpha=0.5) + 
    facet_wrap(~Diagnosis, scale="free") 
pp
pp + facet_grid(Diagnosis~factor(Time_seq),scale="free")

setkey(meta, "SeqID")
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), HB_Mayo_impu <= 5],
                geom.ind = c("point"))
fviz_pca_biplot(pca,
                select.var=list( contrib=30),
                repel=TRUE,
                habillage = meta[rownames(pca[["x"]]), Response],
                geom.ind = c("point"))

pca_d <- data.table(reshape2::melt(pca[["x"]][,1:10], varnames=c("SeqID","PC")))
pca_d <- merge(pca_d,meta, by = "SeqID")

pp <- ggplot(pca_d, aes(y=value, x = HB_Mayo_impu)) +
    geom_point() +
    geom_smooth(method="loess", k = 10) +
    facet_wrap(PC~Diagnosis)
pp
```

Thats rather shitty - compared to the results obtained from the FVA.


# FVA range reaction to predict response

```{r multinomialTests}

for(tm in c(0,14)){
multnom_dat <- fva[,meta[Time_seq == tm & Tissue == "Biopsy", SeqID], "range"]
multnom_dat <- multnom_dat[apply(multnom_dat,1,var) > 1E-6,]
multnom_dat <- reshape2::melt(multnom_dat)
multnom_dat <- merge(data.table(multnom_dat),meta, by.x="sample",by.y="SeqID")
multnom_dat[,Response := relevel(factor(Response), "R")]
multnom_dat[,Diagnosis := factor(Diagnosis)]

models <- list()
stats <- list()
dmp <- capture.output(mod0 <- multinom(data = meta[Time_seq ==tm & Tissue == "Biopsy",], Response ~ Diagnosis))
for(reac in unique(multnom_dat[,rxn])){
    dat <- multnom_dat[rxn == reac,]
    dmp <- capture.output(mod <- multinom(data = dat, Response~Diagnosis*value))
    LLR <- anova(mod0,mod)[["Pr(Chi)"]][2]
    if(LLR < 0.05){
        models[[reac]] <- mod
        ano <- Anova(mod, type = 3)
        sumMod <- summary(mod)
        z <- sumMod$coefficients/sumMod$standard.errors
        p <- (1 - pnorm(abs(z), 0, 1)) * 2
        s<- cbind(rxn = reac,t(p)[-1,], as.data.frame(ano))
        s <- cbind(coef = rownames(s), s)
        stats[[reac]] <- s
    }
}

stats <- data.table(do.call(rbind, stats))
stats[,padj := p.adjust(`Pr(>Chisq)`, method = "BH"), by = coef]
stats[,NR_adj := p.adjust(NR, method = "BH"), by = coef]

reac <- stats[padj <= 0.05 & NR_adj <= 0.05, rxn]
if(length(reac) > 0){
DT::datatable(stats[padj <= 0.05 & NR_adj <= 0.05,])

pp <- ggplot(multnom_dat[rxn %in% reac,], aes(x=Response, y = value, fill = Diagnosis)) +
    geom_boxplot() + 
    geom_jitter(height = 0, width = 0.25) +
    facet_wrap(~rxn, scale ="free") +
    ggtitle(paste0(tm," days after treatment start"))
print(pp)

lapply(models[reac],function(x) exp(coef(x)))
} else {
    print("No significant reactions")
}
}

```

