---
title: Statistical analysis of the FBA for IBD-patient specific models
author: Jan Taubenheim
date: "`r Sys.Date()`"
output: 
        html_document:
                toc: true
                theme: united
---

```{r loadData, eval = TRUE, include=FALSE}

log <- file(snakemake@log[[1]], open = "wt")
sink(log)
sink(log, type = "message")

knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.width = 10, fig.height=5)

require(data.table)
require(ggplot2)
#require(lqmm)
require(doMC)
require(parallel)
require(foreach)
require(lme4)
require(lmerTest)


```


```{r debug, eval=FALSE, include = FALSE}

# This section is only for debugging purposes
setClass("snakemake", representation(input = "list", output = "list"))
snakemake <- new("snakemake", 
                    input = list(
                        fba_results = "results/data/FBA/MatjesAbsorption.colormore22-fluxes.csv",
                        subSysCount = "results/data/ModelAnalysis/MatjesAbsorption.colormore22-SubSysMat.csv",
                        meta = "resources/META_emed_future.csv",
                        clinic = "resources/ClinicalData.csv",
                        subsysTable = "resources/Subsystems.csv",
                        sampleBlacklist = "resources/sampleBlacklist.csv"
                    ),
                    output = list(
                        html = "results/reports/MatjesAbsorption.colormore22-Natrix_analyseRxnCounts.html"
                    )
                )

```

```{r loadData2}

# meta data
meta <- fread(snakemake@input[["meta"]])
meta[,PatientID_Time := paste(PatientID, Time_seq, sep ="_")]
meta <- meta[Time_seq != -1,]
clinic <- fread(snakemake@input[["clinic"]], drop=1)
meta <- merge(meta,clinic, by="PatientID_Time", all.x=TRUE, suffix = c(".meta",""))

# remove degraded samples
blcklst <- fread(snakemake@input[["sampleBlacklist"]])
blckSmpls <- blcklst[relevance < 2, Sample]
meta <- meta[!(SeqID %in% blckSmpls),]

# FBA data
fba <- read.csv(snakemake@input[["fba_results"]],row.names=1)
meta <- meta[SeqID %in% colnames(fba),]

```

The general idea is to see patterns in the fluxes of the patient metabolism to associate that back to the HB/Mayo score or the remission state of the patients.We are generally interested in the Biopsy samples only, but blood might become interesting at another point in the project, thus is was processed in the same manner.

# PCA 

To get an impression of the data lets perform a PCA.

```{r PCA}

fba_biopsy <- fba[,unlist(meta[Tissue=="Biopsy",SeqID])]
pca <-prcomp(fba_biopsy, scale = TRUE, center =TRUE)
pca_dat <- merge(meta, data.table(pca$rotation, keep.rownames=TRUE), by.x="SeqID", by.y="rn")
p <- ggplot(pca_dat, aes(x=PC1,y=PC2, color = HB_Mayo_impu < 5, shape =Diagnosis)) +
    geom_point(size=3,alpha=0.5)
p
p <- ggplot(pca_dat, aes(x=PC1,y=PC2, color = log2(Time_seq), shape =Cohort)) +
    geom_point(size=3,alpha=0.5)
p
p <- ggplot(pca_dat, aes(x=PC1,y=PC2, color = Treatment, shape =Remission)) +
    geom_point(size=3,alpha=0.5)
p

p <- ggplot(pca_dat, aes(x=PC3,y=PC1, color = HB_Mayo_impu < 5, shape =Diagnosis)) +
    geom_point(size=3,alpha=0.5)
p

```

The PCA looks weird, but there seems to be a good fraction of highly inflamed people which can be discriminated from the rest. This appears similar to the FVA and reaction count results, though the "rest" is more clustered together here.

# Objective function

Lets plot the results of the objective function in relation to the HB/Mayo score.

```{r}

biomass <- grep("biomass", rownames(fba), ignore.case=TRUE, value=TRUE)
biomass_dat <- reshape2::melt(t(fba[biomass,]), varnames=c("SeqID","Rxn"), value.name="Flux")
biomass_dat <- merge(meta, data.table(biomass_dat), by = "SeqID")

pp <- ggplot(biomass_dat, aes(x=HB_Mayo_impu, y = Flux, color = log2(Time_seq))) +
         geom_point() +
         facet_wrap(~Rxn, scale = "free_y")
pp

pp <- ggplot(biomass_dat[Tissue=="Biopsy",], aes(x=HB_Mayo_impu, y = Flux, color = log2(Time_seq))) +
         geom_point() +
         facet_wrap(~Rxn, scale = "free_y")+
         ggtitle("Biopsy")
pp

```

Ok nothing obvious to see here. Lets look for statistical interesting reactions, which are associated with HB/Mayo scores.

# LQMM
The calculations take really long, thus I did not include it for now, as it is only a side results.

```{r lqmm, eval=FALSE, include = FALSE}

lqmm_dat <- merge(meta[Tissue == "Biopsy",],
                  data.table(t(fba), keep.rownames = TRUE),
                  by.x = "SeqID",
                  by.y = "rn")

getVariablesFromFormula <- function(RHS_form){
    # the function basically splits the LHS of the formula for adonis into the
    # final coeffecients which are calculated - this is needed to construct the
    # summary table
    facs <- gsub("^~| ","", RHS_form)
    facs <- strsplit(facs, split="\\+")[[1]]
    facs <- sapply(facs, gsub, pattern ="(.*)\\*(.*)", replacement = "\\1\\+\\2\\+\\1:\\2")
    facs <- unlist(strsplit(facs, split="\\+"))
    return(facs)
}

# short function to do the actual test
lqmmShort <- function(dat, formula_LHS, formula_RHS, random = as.formula("~1"), group="PatientID"){
    require(data.table)
   # 
    dat <- data.table(dat)
    # first check the formula and what to extract
    facs <- getVariablesFromFormula(formula_RHS)
    if(class(random) == "character"){
        random <- as.formula(random)
    }
    # hack to make the lqmm work in an automated fashion
    dat[,group := get(group)]
    dat[,rxn := get(formula_LHS)]
#
    fom <- paste0("rxn","~",formula_RHS)
    mod <- lqmm(data = dat,
            fixed = as.formula(fom),
            random = random,
            group = group)

    # another hack to get the correct formula into the summary function
    Call <- match.call()
    dd <- Call[["dat"]]
    mod$call <- str2lang(paste0("lqmm(data = ",dd, ", fixed = ", fom, ", random = ", paste(random, collapse = ""),", group = ",group,")"))
    sum_mod <- summary(mod)
    res <- data.frame(formula_LHS = formula_LHS,
                formula_RHS = formula_RHS,
                variable = facs,
                coefficient = sum_mod[["tTable"]][facs,1],
                stdErr = sum_mod[["tTable"]][facs,2],
                lw_bnd = sum_mod[["tTable"]][facs,3],
                up_bnd = sum_mod[["tTable"]][facs,4],
                pval = sum_mod[["tTable"]][facs,5],
                random = paste(as.character(random),collapse = ""),
                group = group
    )
    return(res)
}

registerDoMC(cores = detectCores())
require(tictoc)
tic()
results <- foreach(rxn = rownames(fba)[1:100], .combine="rbind") %dopar%
    lqmmShort(dat = lqmm_dat,
              formula_LHS = rxn,
              formula_RHS = "HB_Mayo_impu",
              random = as.formula("~1"),
              group="PatientID")
toc()

```

However, we can perform a LMM for each reaction and assume normality.

```{r LMM}

fba_clean <- fba[apply(fba,1,var) > 1E-6,]
ori_names <- rownames(fba_clean)
rownames(fba_clean) <- 
    paste0("R_",gsub("\\((.*)\\)","_LPAREN_\\1_RPAREN_",rownames(fba_clean)))
lmm_dat <- merge(meta[Tissue == "Biopsy",],
                  data.table(t(fba_clean), keep.rownames = TRUE),
                  by.x = "SeqID",
                  by.y = "rn")

registerDoMC(cores = detectCores() -1)
mods <- foreach(rxn = rownames(fba_clean)) %dopar%
    lmer(data=lmm_dat,
         formula = as.formula(paste("HB_Mayo_impu ~",rxn,"+ (1|PatientID)")))
names(mods) <- ori_names
coefs <- lapply(mods, function(x) coefficients(summary(x)))
for(n in names(coefs)){
    coefs[[n]] <- cbind(Rxn=n, coefficient = rownames(coefs[[n]]), data.frame(coefs[[n]]))
}
coef_tab <- do.call(rbind, coefs)
colnames(coef_tab) <- c("Rxn","Coefficient","Estimate","StdError","DF", "tValue","pval")
coef_tab <- data.table(coef_tab)
coef_tab <- coef_tab[Coefficient != "(Intercept)",]
coef_tab[,padj := p.adjust(pval, method = "BH")]

coef_tab[,Rxn := factor(Rxn, levels = Rxn[order(Estimate)])]
p <- ggplot(data = coef_tab[padj <= 0.05,],
            aes(x = Rxn, y= Estimate, color = padj)) +
            geom_point() +
            geom_errorbar(aes(ymax = StdError+Estimate, ymin = Estimate - StdError)) +
            theme(axis.text.x = element_text(angle=45, vjust = 1, hjust=1))
p


```

Interesting, but not more...

