---
title: Statistical analysis of the count matrix of extracted reactions
author: Jan Taubenheim
date: "`r Sys.Date()`"
output: 
        html_document:
                toc: true
                theme: united
---

# Synopsis

To get an idea about the metabolic capabilities of the extracted models.

```{r loadLibraries}

log <- file(snakemake@log[[1]], open = "wt")
sink(log)
sink(log, type = "message")

knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.width = 10, fig.height=5)

require(ggplot2)
require(data.table)
require(lme4)
if(!require(lqmm)){
   install.packages("lqmm",repos="https://cloud.r-project.org")
   require(lqmm)
}

require(car)
require(elasticnet)
require(FactoMineR)
require(factoextra)
require(psych)
require(ggpubr)
require(cowplot)
require(DT)
require(ggridges)
require(pheatmap)
require(limma)
require(clusterProfiler)

```

```{r debug, eval=FALSE, include = FALSE}

# This section is only for debugging purposes
setClass("snakemake", representation(input = "list", output = "list"))
snakemake <- new("snakemake", 
                    input = list(
                        rxnCount = "results/data/ModelAnalysis/MatjesAbsorption.colormore22-rxnIdMat.csv",
                        subSysCount = "results/data/ModelAnalysis/MatjesAbsorption.colormore22-SubSysMat.csv",
                        meta = "resources/META_emed_future.csv",
                        clinic = "resources/ClinicalData.csv",
                        subsysTable = "resources/Subsystems.csv",
                        sampleBlacklist = "resources/sampleBlacklist.csv"
                    ),
                    output = list(
                        html = "results/reports/MatjesAbsorption.colormore22-Natrix_analyseRxnCounts.html"
                    )
                )

```

```{r loadDataSet}

meta <- fread(snakemake@input[["meta"]])
meta[,PatientID_Time := paste(PatientID, Time_seq, sep ="_")]
meta <- meta[Time_seq != -1,]
clinic <- fread(snakemake@input[["clinic"]], drop=1)
meta <- merge(meta,clinic, by="PatientID_Time", all.x=TRUE, suffix = c(".meta",""))
#meta_merge <- merge(meta,clinic, all.x=TRUE)

subSys <- fread(snakemake@input[["subSysCount"]])
subSysCount <- as.matrix(subSys[,-1, with =F])
rownames(subSysCount) <- gsub(" ","_",unlist(subSys[,1]))
rxns <- fread(snakemake@input[["rxnCount"]])
rxnCount <- as.matrix(rxns[,-1,with=F ])
rownames(rxnCount) <- gsub(" ","_",unlist(rxns[,1]))
subSystems <- fread(snakemake@input[["subsysTable"]])

# remove degraded samples
blcklst <- fread(snakemake@input[["sampleBlacklist"]])
blckSmpls <- blcklst[relevance < 2, Sample]

meta <- meta[!(SeqID %in% blckSmpls),]
subSysCount <- subSysCount[,!(colnames(rxnCount) %in% blckSmpls)]
rxnCount <- rxnCount[,!(colnames(rxnCount) %in% blckSmpls)]
meta <- meta[SeqID %in% colnames(rxnCount),]

```


# Analysis

This part will analysis the data.

## General statistics

Plot a PCA of the Jaccard distance of the reactions presence in the data set

```{r PCAJaccard}

distJac <- dist(t(rxnCount),method = "binary")
jac_pca <- prcomp(distJac)

explVar <- round(jac_pca$sdev^2/sum(jac_pca$sdev^2) * 100, 2)
jacPCA_dat <- data.table(jac_pca$x, keep.rownames=TRUE)
jacPCA_dat <- merge(jacPCA_dat, meta, by.x ="rn",by.y="SeqID")



pcaPlot <- ggplot(jacPCA_dat, aes(x=PC1,y=PC2, color = Cohort, shape = Tissue)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVar[2],"%)")) +
    xlab(paste0("PC1 (",explVar[1],"%)"))
pcaPlot

pcaPlot <- ggplot(jacPCA_dat, aes(x=PC1,y=PC2, color = HB_Mayo_impu <= 5, shape = Tissue)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVar[2],"%)")) +
    xlab(paste0("PC1 (",explVar[1],"%)"))
pcaPlot

pcaPlot <- ggplot(jacPCA_dat, aes(x=PC1,y=PC2, color = Gender, shape = Tissue)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVar[2],"%)")) +
    xlab(paste0("PC1 (",explVar[1],"%)"))
pcaPlot

pcaPlot <- ggplot(jacPCA_dat, aes(x=PC1,y=PC2, color = Diagnosis, shape = Tissue)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVar[2],"%)")) +
    xlab(paste0("PC1 (",explVar[1],"%)"))
pcaPlot


distJacBlood <- dist(t(rxnCount[,meta[Tissue == "Blood", SeqID]]),method = "binary")
jac_pcaBlood <- prcomp(distJacBlood)

explVarBlood <- round(jac_pcaBlood$sdev^2/sum(jac_pcaBlood$sdev^2) * 100, 2)
jacPCABlood_dat <- data.table(jac_pcaBlood$x, keep.rownames=TRUE)
jacPCABlood_dat <- merge(jacPCABlood_dat, meta, by.x ="rn",by.y="SeqID")

pcaPlot_Blood <- ggplot(jacPCABlood_dat, aes(x=PC1,y=PC2, color = Cohort, shape = Gender)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVarBlood[2],"%)")) +
    xlab(paste0("PC1 (",explVarBlood[1],"%)")) +
    ggtitle("Blood")
pcaPlot_Blood 

pcaPlot_Blood <- ggplot(jacPCABlood_dat, aes(x=PC1,y=PC2, color = HB_Mayo_impu <= 5, shape = Gender)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVarBlood[2],"%)")) +
    xlab(paste0("PC1 (",explVarBlood[1],"%)")) +
    ggtitle("Blood")
pcaPlot_Blood 

distJacBiopsy <- dist(t(rxnCount[,meta[Tissue == "Biopsy", SeqID]]),method = "binary")
jac_pcaBiopsy <- prcomp(distJacBiopsy)

explVarBiopsy <- round(jac_pcaBiopsy$sdev^2/sum(jac_pcaBiopsy$sdev^2) * 100, 2)
jacPCABiopsy_dat <- data.table(jac_pcaBiopsy$x, keep.rownames=TRUE)
jacPCABiopsy_dat <- merge(jacPCABiopsy_dat, meta, by.x ="rn",by.y="SeqID")

pcaPlot_Biopsy <- ggplot(jacPCABiopsy_dat, aes(x=PC1,y=PC2, color = Cohort, shape = Gender)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVarBiopsy[2],"%)")) +
    xlab(paste0("PC1 (",explVarBiopsy[1],"%)")) +
    ggtitle("Biopsy")
pcaPlot_Biopsy 

pcaPlot_Biopsy <- ggplot(jacPCABiopsy_dat, aes(x=PC1,y=PC2, color = HB_Mayo_impu <= 5, shape = Gender)) +
    geom_point(size=2) +
    theme_bw() +
    ylab(paste0("PC2 (",explVarBiopsy[2],"%)")) +
    xlab(paste0("PC1 (",explVarBiopsy[1],"%)")) +
    ggtitle("Biopsy")
pcaPlot_Biopsy 

```


## Tryptophan metabolism

Konrad Aden is interested in specifically the tryptophan metabolism of the patients. So lets start with that and look whether differences exist in the remitters and non-remitters group.

```{r plotTryptophan, fig.width = 9, fig.height=5}

tr_id <- grep("Tryptophan", rownames(subSysCount), value=TRUE)
hist(log(subSysCount[tr_id,]+1))

tSubSysCount <- t(subSysCount)
subSysData_data <- base::merge(meta,data.table(tSubSysCount, keep.rownames=TRUE), by.x="SeqID",by.y="rn")

pp <- ggplot(subSysData_data[Tissue=="Biopsy",], aes_string(y=tr_id, fill ="Remission")) + geom_boxplot(aes(x=as.factor(Time_seq)))+
    theme_bw()

pp

```

There seems to be an effect in the patients, lets check it in a model.

```{r lmerTryptophan, fig.width = 9, fig.height = 5}

mod <- lmer(data = subSysData_data[Tissue == "Biopsy",],
            formula = log(HB_Mayo_impu+1) ~
                Tryptophan_metabolism +
                (1|PatientID)
            )
summary(mod)
Anova(mod)
plot(mod)
qqPlot(resid(mod, "pearson"))
infMod <- influence(mod)
plot(cooks.distance(infMod))

pp <- ggplot(subSysData_data[Tissue=="Biopsy",], aes(x=log(HB_Mayo_impu+1), y=Tryptophan_metabolism))+
    geom_point(aes(color=Cohort, shape = Diagnosis)) +
    geom_smooth(method="lm") +
    stat_cor()+
    theme_bw()

pp

```

Well that's highly significant. There is a clear correlation between number of reactions extracted from the tryptophan metabolism and the HB-Mayo scores.

## Checking single LMM for association in Mayo-scores - Inflammation association

Similar to the approach for in DESeq2, one could generate a table of changes between the subsystems and the HB_mayo score.

```{r LinarMixedModels, include = FALSE, eval = FALSE}

tSubSysCount <- t(subSysCount)
subSysData_data <- base::merge(meta,data.table(tSubSysCount, keep.rownames=TRUE), by.x="SeqID",by.y="rn")
subSysBiopsy <- subSysData_data[Tissue == "Biopsy",]
subsystems <- rownames(subSysCount)
LMM_results <- data.frame(Subsystem = subsystems,
                          Coefficient = as.numeric(rep(NA,length(subsystems))),
                          StdError = as.numeric(rep(NA, length(subsystems))),
                          pvalue = as.numeric(rep(NA, length(subsystems))),
                          shapiro_resi = as.numeric(rep(NA, length(subsystems))),
                          KS_resi = as.numeric(rep(NA, length(subsystems))),
                          skew = as.numeric(rep(NA, length(subsystems))),
                          kurt = as.numeric(rep(NA, length(subsystems)))
)
rownames(LMM_results) <- subsystems

for(ss in subsystems){
    d_subsys <- subSysBiopsy[,c("HB_Mayo_impu","PatientID",ss),with=FALSE]
    colnames(d_subsys)[3] <- "subsys"
    if(var(d_subsys[,subsys]) >0){
        mod <- lmer(data = d_subsys,
                    formula = HB_Mayo_impu~
                        subsys +
                        (1|PatientID),
                    )
        res <- c(
                 as.numeric(summary(mod)[["coefficients"]][2,1:2]),
                 Anova(mod)[["Pr(>Chisq)"]],
                 shapiro.test(resid(mod))[["p.value"]],
                 ks.test(resid(mod),"pnorm",mean(resid(mod)),sd(resid(mod)))[["p.value"]],
                 skew(resid(mod)),
                 kurtosi(resid(mod))
                 )

        names(res) <-  colnames(LMM_results)[-1]
        LMM_results[ss,2:ncol(LMM_results)] <- res
    }
}
LMM_results <- data.table(LMM_results)
LMM_results[,padj := p.adjust(pvalue)]

p_LMM_coefficients <- ggplot(LMM_results[padj<=0.05,], aes(x = Subsystem, y = Coefficient, color = padj )) +
    geom_point(size=3) +
    geom_errorbar(aes(ymin = Coefficient-StdError, ymax = Coefficient+as.numeric(StdError)))+
    scale_color_gradient(low="Blue", high ="Red") +
    theme_bw()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust=1))

p_LMM_coefficients

DT::datatable(LMM_results)

```


```{r LQMM}

tSubSysCount <- t(subSysCount)
subSysData_data <- base::merge(meta,data.table(tSubSysCount, keep.rownames=TRUE), by.x="SeqID",by.y="rn")
subSysBiopsy <- subSysData_data[Tissue == "Biopsy",]
subsystems <- rownames(subSysCount)

LQMM_results <- data.frame(Subsystem = subsystems,
                          Coefficient = as.numeric(rep(NA,length(subsystems))),
                          StdError = as.numeric(rep(NA, length(subsystems))),
                          pvalue = as.numeric(rep(NA, length(subsystems))),
                          resiHomoSced = as.numeric(rep(NA,length(subsystems)))
)
rownames(LQMM_results) <- subsystems
LQMM_models <- list()

for(ss in subsystems){
    d_subsys <- subSysBiopsy[,c("HB_Mayo_impu","PatientID",ss),with=FALSE]
    colnames(d_subsys)[3] <- "subsys"
    if(var(d_subsys[,subsys]) >0){
        d_subsys <- d_subsys[apply(d_subsys,1,function(x) !any(is.na(x))),]
        print(ss)
        mod <- lqmm(data = d_subsys,
                    fixed = HB_Mayo_impu ~
                        subsys, 
                    random = ~1,
                    group = "PatientID",
                    type = "robust"
                    )
        LQMM_models[[ss]] <- mod
        summ <- summary(mod, R=100)
        res <- c(
                 summ[["tTable"]][2,c(1:2,5)],
                 cor.test(resid(mod)^2,predict(mod))[["p.value"]]
                 )
        names(res) <-  colnames(LQMM_results)[-1]
        LQMM_results[ss,2:ncol(LQMM_results)] <- res
    }
}

LQMM_results <- data.table(LQMM_results)
LQMM_results[,padj := p.adjust(pvalue)]

p_LQMM_coefficients <- ggplot(LQMM_results[padj<=0.05,], aes(x = Subsystem, y = Coefficient, color = padj )) +
    geom_point(size=3) +
    geom_errorbar(aes(ymin = Coefficient-StdError, ymax = Coefficient+StdError))+
    scale_color_gradient(low="Blue", high ="Red") +
    theme_bw()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust=1)) 
p_LQMM_coefficients

DT::datatable(LQMM_results)
```

The plot shows the coefficients with standard errors for the linear quantile mixed model of HB/Mayo scores to number of reactions in the metabolic pathways.

```{r plotCoeffecientsToData, fig.height= 8}

for(subsys in LQMM_results[padj<=0.05, Subsystem]){
    ss <- gsub("[,|/]",".",subsys)
    mod <- LQMM_models[[subsys]]
    dat <- data.frame(subSysBiopsy)
    colnames(dat) <- gsub("[,|/]",".",colnames(dat))
    dat[["slope"]] <- coef(mod)[2]
    dat[["intercept"]] <- ranef(mod)[dat[["PatientID"]],1] + coef(mod)[1]
    stdErr <- summary(mod)[["tTable"]][,1]
    pp <- ggplot(dat, aes_string(x=ss, y="HB_Mayo_impu")) +
        theme_bw()+
        geom_point() +
        geom_abline(aes(slope = max(slope), intercept=intercept))+
        geom_abline(aes(slope = max(slope)-stdErr[2], intercept=intercept-stdErr[1]), linetype = "dashed")+
        geom_abline(aes(slope = max(slope)+stdErr[2], intercept=intercept+stdErr[1]), linetype = "dashed")+
        facet_wrap(~PatientID) +
        ggtitle(subsys)
    print(pp)
}

```


## Ordination of Subsystems - Inflammation association

Lets look at other correlations in the data.

```{r, correlations, fig.width = 9, fig.height=5, warnings =FALSE, messages = FALSE}

tSubSysCount <- t(subSysCount)
subSysData_data <- base::merge(meta,data.table(tSubSysCount, keep.rownames=TRUE), by.x="SeqID",by.y="rn")
subSysData_data[,logMayo := log(HB_Mayo_impu+1)]
subSysData_data <- subSysData_data[Tissue =="Biopsy" & !is.na(HB_Mayo_impu),]

# test correlation between counts and HB mayo
corr <- c(rep(0, nrow(subSysCount)))
pvals<-c(rep(1, nrow(subSysCount)))
for(i in 1:nrow(subSysCount)){
    subsys <- rownames(subSysCount)[i]
    if(var(subSysData_data[, subsys, with =FALSE]) != 0){
        corrTest <- cor.test(unlist(subSysData_data[,HB_Mayo_impu]), 
                             as.vector(unlist(subSysData_data[,subsys,with=F])),
                             method = "spearman")
        corr[i] <- corrTest[["estimate"]]
        pvals[i] <- corrTest[["p.value"]]
    } 
}
corrTest <- data.table(subSystem = rownames(subSysCount),
                       correlation = corr,
                       pvalue = pvals,
                       padj = p.adjust(pvals, method ="BH"))

pairs.panels(subSysData_data[Tissue=="Biopsy",c("HB_Mayo_impu",unlist(corrTest[padj <= 0.05 & abs(correlation) >= 0.3,subSystem])), with =FALSE], stars=TRUE, method ="spearman",main="Spearman, |rho| >= 0.3")

subSysCount_long <- reshape2::melt(subSysCount)
colnames(subSysCount_long) <- c("Subsystem","Sample","value")
subSysCount_long <- merge(data.table(subSysCount_long), meta, by.x ="Sample", by.y = "SeqID")

pp_corr <- ggplot(data=subSysCount_long[Subsystem %in% corrTest[padj <= 0.05, subSystem],], aes(x = HB_Mayo_impu, y =value)) + 
    geom_point() +
    geom_smooth(method ="lm") +
    facet_wrap(~Subsystem, scale = "free_y") +
    theme_bw()
pp_corr

```

There are in total `r sum(corrTest[,padj] <= 0.05)`(`r nrow(corrTest)`)  subsystems significantly correlated with the HB_Mayo_impu score. To reduce correlation of predictors, one can calculate principal components and use these as measure for prediction.

```{r PCA, fig.width = 9, fig.height=5}

biopsy <- meta[Tissue== "Biopsy" & SeqID %in% colnames(subSysCount) & !is.na(HB_Mayo_impu), SeqID]
biopsy <- tSubSysCount[biopsy,]
biopsy <- biopsy[,apply(biopsy,2,var) >0]
PCA <- prcomp(biopsy, scale=TRUE, center=TRUE)
fviz_eig(PCA)

pca_data <- merge(meta, data.table(PCA$x, keep.rownames=TRUE), by.x="SeqID",by.y="rn")
expl_var <- PCA$sdev^2/sum(PCA$sdev^2)*100


pp <- ggplot(pca_data[Tissue=="Biopsy",], aes(x=PC1, y=PC2, color =HB_Mayo_impu<=5, shape=Diagnosis, label = Time_seq)) +
    geom_text()+
    ylab(paste0("PC2 (",round(expl_var[2],digits=2),"%)")) +
    xlab(paste0("PC1 (",round(expl_var[1],digits=2),"%)")) +
    theme_bw()
pp

pp <- ggplot(pca_data[Tissue=="Biopsy",], aes(x=PC1, y=PC2, color =Remission, shape=Diagnosis, label = Time_seq)) +
    geom_text()+
    ylab(paste0("PC2 (",round(expl_var[2],digits=2),"%)")) +
    xlab(paste0("PC1 (",round(expl_var[1],digits=2),"%)")) +
    theme_bw()

pp

pp <- ggplot(pca_data[Tissue=="Biopsy",], aes(x=PC1, y=PC2, color =Time_seq, shape=Remission, label = Time_seq)) +
    geom_point(size=3)+
    ylab(paste0("PC2 (",round(expl_var[2],digits=2),"%)")) +
    xlab(paste0("PC1 (",round(expl_var[1],digits=2),"%)")) +
    theme_bw()
pp



sel <- c("HB_Mayo_impu",paste0("PC",1:10))
pairs.panels(pca_data[Tissue=="Biopsy",sel, with =FALSE], stars=TRUE, method="spearman", main ="Spearman correlation")

# calculate the correlation for each PC and the HB_Mayo_impu 
pcs <- grep("PC", colnames(pca_data), value =TRUE)
corr <- rep(0, length(pcs))
pvals<-rep(1, length(pcs))
for(i in 1:length(pcs)){
    pc <- pcs[i]
    if(var(pca_data[, pc, with =FALSE]) != 0){
        corrTest <- cor.test(unlist(pca_data[,HB_Mayo_impu]), unlist(pca_data[,pc, with =FALSE]), method ="spearman")
        corr[i] <- corrTest[["estimate"]]
        pvals[i] <- corrTest[["p.value"]]
    } 
}
corrTest <- data.table(PC = pcs,
                       correlation = corr,
                       pvalue = pvals,
                       padj = p.adjust(pvals, method ="BH"),
                       axis = as.integer(gsub("PC","",pcs)))

# get contributions for all significant correlations

fviz_contrib(PCA, choice = "var", axes = corrTest[padj <= 0.05, axis])

plots_var <- list()
plots_ind <- list()
for(pc in corrTest[padj <= 0.05, axis]){
    plots_var[[paste0("PC",pc)]] <- fviz_contrib(PCA, choice = "var", axes = pc, top=10)
    plots_ind[[paste0("PC",pc)]] <- fviz_contrib(PCA, choice = "ind", axes = pc, top=10)
}
pp <- plot_grid(plotlist = plots_var)
pp
pp <- plot_grid(plotlist = plots_ind)
pp
```

The variability captured on the `r corrTest[padj <= 0.05, PC]` is/are correlated with the variation in the HB_Mayo_impu score.

```{r PCAindividualContribution,fig.width = 9, fig.height=5}

pca_contrib <- get_pca_ind(PCA)
pca_indContr <- merge(meta,
                      data.table(pca_contrib[["contrib"]], keep.rownames =TRUE),
                      by.x = "SeqID",
                      by.y = "rn")
pca_indContr <- melt(pca_indContr, measure.vars = colnames(pca_contrib[["contrib"]]), variable.name = "PrincipalComponent", value.name = "Contribution")
# extract only the correlated components
pca_indContr <- pca_indContr[PrincipalComponent %in% paste0("Dim.",corrTest[padj <= 0.05, axis]),]

# define factors to check
facs <- c("Gender","Response","Remission","Therapy", "Diagnosis","Cohort")

plots <- list()
for(vr in facs){
    dtbl <- pca_indContr[,.(Contribution = sum(Contribution)), by =c(vr, "PrincipalComponent")]
    ratio <- pca_indContr[,.(PrincipalComponent = "smpl_ratio", Contribution = .N/nrow(pca_indContr)*100), by = vr]
    dtbl <- rbind(ratio, dtbl)
    plots[[vr]] <- ggplot(dtbl, aes_string(x="PrincipalComponent", y="Contribution", fill = vr)) +
        geom_bar(stat="identity") +
        theme_bw()+  
        theme(axis.text.x= element_text(angle = 90, vjust=0.5, hjust=1))

}

pp <- plot_grid(plotlist=plots)
pp

```


After having defined the axis, which capture the variability to explain at least some extend of the inflammation score, one can try to select relevant features using a sparse PCA.

```{r spcaInflammation, eval = FALSE, include = FALSE}

# define a small function to summarize spca a little
summarizeReduction <- function(spca_result, mat, dims = c(1,2)){
    require(FactoMineR)
    dimRetained <- apply(spca_result[["loadings"]][,dims],2,function(x) sum(x != 0))
    dimReduction <- 1-dimRetained/nrow(spca_result[["loadings"]])
    lambda_1 <- spca_result[["para"]][dims]
    lambda <- spca_result[["lambda"]]
    varRetained <- spca_result[["pev"]][dims]
    pca <- PCA(mat, graph = FALSE)
    varRetained_rel <- varRetained/(pca[["eig"]][dims,2]/100)
    df <- data.frame(PCs = dims,
                     lambda = lambda,
                     lambda_1 = lambda_1,
                     dimRetained = dimRetained,
                     dimReduction = dimReduction,
                     varRetained = varRetained,
                     varRetained_rel = varRetained_rel)

    return(df)
}

# number of pcas to keep
dims <- corrTest[padj <= 0.05, axis]
ncp <- max(dims)
dims <- 1:ncp
pca <- PCA(biopsy, graph=FALSE, ncp = ncp)
expVar <- pca$eig[dims,2]/100
biopsy_sc <- scale(biopsy)
lambda  <- 0
# do spca, find upper and lower bounds for lambda_1
lambda_1 = 1
sparse <- spca(biopsy_sc, K = ncp,
           para =lambda_1 * (1-expVar),
            lambda = lambda)


prmSearchSPCA <- summarizeReduction(sparse, biopsy_sc, dims= dims)
dimRetained <- prmSearchSPCA[["dimRetained"]]
dimReduction <- prmSearchSPCA[["dimReduction"]]
while(max(dimRetained > 0)){
    lambda_1 <- lambda_1*10
    sparse <- spca(biopsy_sc, K = ncp,
               para =lambda_1 * (1-expVar),
               lambda = lambda)
    current <- summarizeReduction(sparse, biopsy_sc, dims = dims)
    dimRetained <- current[["dimRetained"]]
    prmSearchSPCA <- rbind(prmSearchSPCA, current)
}

lambda_1 <- 1
while(min(dimReduction) > 0){
    lambda_1 <- lambda_1 *0.1
    sparse <- spca(biopsy_sc, K = ncp,
               para =lambda_1 * (1-expVar),
               lambda = lambda)
    current <- summarizeReduction(sparse, biopsy_sc, dims = dims)
    dimReduction <- current[["dimReduction"]]
    prmSearchSPCA <- rbind(prmSearchSPCA, current)
}

if(sum(prmSearchSPCA[["PCs"]] == 1) < 10){
    minMaxL <- range(prmSearchSPCA[["lambda_1"]])
    for(lambda_1 in 2^seq(log2(minMaxL[1]),log2(minMaxL[2]),length=10)[c(-1,-10)]){
        sparse <- spca(biopsy_sc, K = ncp,
               para =lambda_1 * (1-expVar),
               lambda = lambda)
        current <- summarizeReduction(sparse, biopsy_sc, dims = dims)
        prmSearchSPCA <- rbind(prmSearchSPCA, current)
    }
}


prmSearchSPCA <- data.table(prmSearchSPCA)
prmSearchSPCA[,score := (dimReduction+varRetained_rel)/2]
lambda_1 <- c()
for(pc in unique(prmSearchSPCA[,PCs])){
    l <- prmSearchSPCA[PCs == pc,]
    lambda_1 <- c(lambda_1, l[score == max(score),lambda_1])
}

lambda_df <- data.frame(lambda_1 = lambda_1, PCs = 1:length(lambda_1))
p_dimReduction <- ggplot(prmSearchSPCA, aes(x= log2(lambda_1), y = dimReduction, color = factor(PCs))) +
    geom_line() +
    geom_point() +
    geom_vline(data=lambda_df, aes(xintercept=log2(lambda_1), color = factor(PCs)))+
    theme_bw()

p_retainedVar <- ggplot(prmSearchSPCA, aes(x = log2(lambda_1), y = varRetained_rel, color = factor(PCs))) +
    geom_line() +
    geom_point()+
    geom_vline(data=lambda_df, aes(xintercept=log2(lambda_1), color = factor(PCs)))+
    theme_bw()

plot_grid(p_retainedVar,p_dimReduction)

# calculate the final spca
sparse <- spca(biopsy_sc, K = ncp, para = lambda_1)
# select only the sparse subsystems to repeat the pca
sparseFacs <- apply(sparse[["loadings"]][,corrTest[padj <= 0.05, axis]], 1,
                    function(x){sum(abs(x))}) != 0
sparseFacs <- names(sparseFacs)[sparseFacs]
sparseFacs

axs <- corrTest[padj <= 0.05, PC]
corr <- corrTest[padj <= 0.05, correlation]
names(corr) <- axs
eig <- (PCA$sdev^2) [corrTest[padj <= 0.05, axis]]
corrEig <- eig*abs(corr)
loads <- sparse[["loadings"]][,axs]

loads_contrib <-apply(loads,2,function(x){abs(x)/sum(abs(x))})
loads_weighted<- t(t(loads_contrib)*corrEig)

loads_wc <- data.frame(weightContrib = apply(abs(loads_weighted),1,sum)/sum(corrEig))
loads_wc[["Subsystems"]] <- factor(rownames(loads_wc), 
                                   levels = rownames(loads_wc)[order(loads_wc[["weightContrib"]],
                                                   decreasing = TRUE)])

p_contrib <- ggplot(loads_wc[loads_wc$weightContrib != 0,],
                    aes(x=Subsystems, y = weightContrib)) +
    geom_bar(stat="identity", fill= "dark blue") +
    ggtitle(paste0("Sparse (relevant) contributions to PC",paste(corrTest[padj <=0.05, axis], collapse = "-"))) +
    ylab("Rho-weighted contributions") +
    xlab("Metabolic subsystem") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1,vjust=1))
p_contrib
```

The last plot shows the most relevant contributions to the dimensions weighted by the correlation with the HB-Mayo score.


```{r Test, include = FALSE, echo = FALSE, eval = FALSE}

sel <-apply(sparse[["loadings"]], 1, function(x){sum(abs(x))}) != 0
biopsy_sparse <- biopsy[,sel]
pca_sparse <- PCA(biopsy_sparse, graph =FALSE, ncp = ncol(biopsy_sparse))
setkey(meta,"SeqID")

fviz_pca_ind(pca_sparse,
             habillage = meta[rownames(biopsy_sparse), HB_Mayo_impu <= 3],
             palette = scales::hue_pal()(2),
             axes = c(1,2),
             geom = "point",
             #select.ind = list(name = unlist(meta[rownames(biopsy_sparse),][Remission != "NR","SeqID"]),
             #select.var = list(contrib = 20),
             repel = TRUE
)

fviz_pca_ind(pca_sparse,
             habillage = factor(meta[rownames(biopsy_sparse), Remission]),
             palette = scales::hue_pal()(3),
             axes = c(1,2),
             geom = "point"
)


pca_data <- merge(data.table(pca_sparse[["ind"]][["coord"]],keep.rownames=TRUE), meta, by.x = "rn", by.y ="SeqID")
pcs <- grep("Dim", colnames(pca_data), value =TRUE)
corr <- rep(0, length(pcs))
pvals<-rep(1, length(pcs))
for(i in 1:length(pcs)){
    pc <- pcs[i]
    if(var(pca_data[, pc, with =FALSE]) != 0){
        corrTest_2 <- cor.test(unlist(pca_data[,HB_Mayo_impu]), unlist(pca_data[,pc, with =FALSE]), method ="spearman")
        corr[i] <- corrTest_2[["estimate"]]
        pvals[i] <- corrTest_2[["p.value"]]
    } 
}
corrTest_2 <- data.table(PC = pcs,
                       correlation = corr,
                       pvalue = pvals,
                       padj = p.adjust(pvals, method ="BH"),
                       axis = as.integer(gsub("Dim.","",pcs)))

fviz_cos2(pca_sparse, choice = "var",axes=c(1,2,13))
fviz_contrib(pca_sparse, choice = "var",axes=c(1,2,13))

```


## Correspondence analysis - Inflammation

A correspondence analysis is looking at factorial data. Thus I will convert the meta data and the presence-absence matrix for the reactions into contingency tables. We are interested in factors which are associated with higher/lower inflammation, which is measured as HB/Mayo. Thus I will bin the HB/Mayo into some few categories and will table the reactions extracted in the samples.

```{r CA_HBmayo, fig.width =9, fig.height = 5}

trxnCount <- data.table(t(rxnCount), keep.rownames=TRUE)
trxnCount <- melt(trxnCount, id.vars = "rn")
rxn_data <- merge(meta[Tissue == "Biopsy" & !is.na(HB_Mayo_impu),], trxnCount, by.x ="SeqID",by.y="rn")
rxn_data <- rxn_data[,.(count = sum(value)),by=c("variable","HB_Mayo_impu")]
rxn_data <-  dcast(data = rxn_data, variable~HB_Mayo_impu)
rn <- rxn_data[,variable]
rxn_data <- as.matrix(rxn_data[,-1])
rownames(rxn_data) <- rn
rxn_data <- rxn_data[apply(rxn_data, 1, sum) > 0,]


ca <- CA(rxn_data, ncp = 20, graph=FALSE)

fviz_screeplot(ca)
fviz_ca_col(ca, axes=c(1,5))
fviz_ca_row(ca, select.row = list(cos2 = 0.90))
fviz_cos2(ca, choice = "row", top = 60, axes=1:2)
fviz_contrib(ca, choice = "row", top = 60, axes=1)
fviz_contrib(ca, choice = "row", top = 60, axes=2)

# do the preprocessing manually
Prp <- rxn_data/sum(rxn_data)
E <- rowSums(Prp) %o% colSums(Prp)
Z <- (Prp-E)/sqrt(E)

# Z is now a scaled residual matrix, which is usually used in an SVD to generate the PCs in a CA. However, since PCA and SVD are related, it is possible to do a PCA here and achieve similar results. At the same time we can apply sparse PCA to do a feature selection an get only some features of the PCs which explain HB-Mayo scores. 
pca <- PCA(t(Z), ncp = nrow(t(Z)), graph = FALSE)

# lets do an correlation test select those PCs associated with changes in HB_Mayo
corr <- c()
pvals <- c()
for(pc_ in colnames(pca[["ind"]][["coord"]])){
    pc <- pca[["ind"]][["coord"]][,pc_]
    hb <-as.integer(rownames(pca[["ind"]][["coord"]]))
    ct <- cor.test(hb, pc, method = "spearman")
    corr <- c(corr, ct[["estimate"]])
    pvals <- c(pvals, ct[["p.value"]])
}
corrTest <- data.table(PC = colnames(pca[["ind"]]),
                       correlation = corr,
                       pvalue = pvals,
                       padj = p.adjust(pvals, method ="BH"))
```

The CA found `r nrow(corrTest[padj <= 0.05,])` significant associated PCs with HB_mayo.

```{r rxnAssociation}

# than we can still use a simple wilcox test for each reaction (present/absence) to test association with HB_Mayo

trxnCount <- data.table(t(rxnCount), keep.rownames=TRUE)
rxn_wilcox <- merge(meta[Tissue == "Biopsy" & !is.na(HB_Mayo_impu),], 
                    trxnCount, 
                    by.x ="SeqID",
                    by.y="rn")

rxn <- rownames(rxnCount)[apply(rxnCount,1,sum)>0]
wcxTest <- data.frame(
                    log2FC = rep(NA, length(rxn)),
                    pval = rep(NA,length(rxn)),
                    padj = rep(NA,length(rxn)),
                    row.names = rxn)
for(r in rxn){
    sel <- rxn_wilcox[[r]]
    if(sum(sel) < length(sel)-3 & sum(sel) > 3){
        wcx <- wilcox.test(rxn_wilcox[sel, HB_Mayo_impu],
                           rxn_wilcox[!sel,HB_Mayo_impu])
        wcxTest[r,"pval"] <- wcx[["p.value"]]
        wcxTest[r,"log2FC"] <- mean(log2(rxn_wilcox[sel, HB_Mayo_impu]+1)) -
                           mean(log2(rxn_wilcox[!sel,HB_Mayo_impu]+1))
                    }
}
wcxTest[["padj"]] <- p.adjust(wcxTest[["pval"]], method = "BH")
wcxTest <- data.table(wcxTest,keep.rownames=TRUE)
wcxTest[,absLog2FC := abs(log2FC)]
wcxTest[,rxn_plot := factor(rn, level = rn[order(log2FC)])]

pp_wilcox <- ggplot(wcxTest[padj <=0.05,], aes(x=rxn_plot, y=log2FC, fill = log2FC)) +
    theme_bw()+
    geom_bar(stat="identity") +
    scale_fill_gradient2(low="blue",mid="white",high="red",midpoint=0) +
    theme(axis.text.x = element_blank(),
          panel.grid.major= element_blank(),
          panel.grid.minor = element_blank()) +
    ylab("HB/Mayo log2-foldchange") +
    xlab("Significantly associated reaction")
pp_wilcox

sigRxn <- rxnCount[wcxTest[padj <= 0.05,rn],meta[Tissue=="Biopsy",SeqID]]
heatMat <- apply(sigRxn,2,as.integer)
colnames(heatMat) <- colnames(sigRxn)
rownames(heatMat) <- rownames(sigRxn)
col_anno <-as.data.frame(meta[,.(Cohort,
                                 Log10_HB_Mayo = log10(HB_Mayo_impu+1),
                                 Remission,
                                 Log10_Time = log10(Time_seq+1))])
rownames(col_anno) <- meta[,SeqID]

pheatmap(heatMat,
         annotation_col= col_anno,
         #annotation_row = rxn_anno,
         show_colnames = FALSE,
         clustering_distance_rows = "binary",
         clustering_distance_cols = "binary",
         show_rownames = FALSE
)


```

There are `r nrow(wcxTest[padj<= 0.05,])` reactions whose presence/absence is associated with a change in HB/Mayo scores. For details in see the table below:
`r DT::datatable(wcxTest[padj <= 0.05,])`.

```{r enrichSubsystems, eval = FALSE, include=FALSE}
# look for subsystem enrichment
enr <- enricher(wcxTest[padj <= 0.05, rn], TERM2GENE = subSystems)
dotplot(enr) + xlab("Fraction (sig. rxns in subsystem)/(all sig. rxns)")

enr_res <- data.table(enr@result)
enr_res <- enr_res[p.adjust <= 0.05,]
ridge_data <- merge(wcxTest,subSystems, by.x="rn",by.y="rxns")
ridge_data <- merge(ridge_data, enr_res, by.x = "subSys", by.y="ID")
p_ridge <- ggplot(ridge_data[padj<=0.05,], aes(x=log2FC, y = subSys, fill = p.adjust))+
    geom_density_ridges() +
    scale_fill_gradient(low = "blue",high="red")+
    theme_bw()

p_ridge
```


## Correspondence analysis - Prediction

A correspondence analysis is looking at factorial data. Thus I will convert the meta data and the presence-absence matrix for the reactions into contingency tables. We are interested in the prediction of Remission state over time. Thus I will combine the time after treatment and the Remission assignment to a new factor and will table the reactions extracted in the samples.

```{r CA, fig.width = 9, fig.height = 5}

trxnCount <- data.table(t(rxnCount), keep.rownames=TRUE)
trxnCount <- melt(trxnCount, id.vars = "rn")
rxn_data <- merge(meta[Tissue == "Biopsy",], trxnCount, by.x ="SeqID",by.y="rn")
rxn_data[,remission_time := paste(Remission, Time_seq, sep ="_")]
rxn_data <- rxn_data[,.(count = sum(value)),by=c("variable","remission_time")]
rxn_data <-  dcast(data = rxn_data, variable~remission_time)
rn <- rxn_data[,variable]
rxn_data <- as.matrix(rxn_data[,-1])
rownames(rxn_data) <- rn
rxn_data <- rxn_data[apply(rxn_data, 1, sum) > 0,]

ca <- CA(rxn_data, ncp = 20, graph=FALSE)
fviz_screeplot(ca)
fviz_ca_col(ca)
fviz_ca_row(ca, select.row = list(cos2 = 0.8))
fviz_cos2(ca, choice = "row", top = 60, axes=1:2)
fviz_contrib(ca, choice = "row", top = 60, axes=1)
fviz_contrib(ca, choice = "row", top = 60, axes=2)


# repeat on the level of cohort to validate the result
# FUTURE only
trxnCount <- data.table(t(rxnCount), keep.rownames=TRUE)
trxnCount <- melt(trxnCount, id.vars = "rn")
rxn_data <- merge(meta[Cohort == "FUTURE" & Tissue == "Biopsy",], trxnCount, by.x ="SeqID",by.y="rn")
rxn_data[,remission_time := paste(Remission, Time_seq, sep ="_")]
rxn_data <- rxn_data[,.(count = sum(value)),by=c("variable","remission_time")]
rxn_data <-  dcast(data = rxn_data, variable~remission_time)
rn <- rxn_data[,variable]
rxn_data <- as.matrix(rxn_data[,-1])
rownames(rxn_data) <- rn
rxn_data <- rxn_data[apply(rxn_data, 1, sum) > 0,]
ca_FUTURE <- CA(rxn_data, ncp = 20, graph=FALSE)
fviz_ca_col(ca_FUTURE)+ ggtitle("FUTURE")

# eMED only
trxnCount <- data.table(t(rxnCount), keep.rownames=TRUE)
trxnCount <- melt(trxnCount, id.vars = "rn")
rxn_data <- merge(meta[Cohort == "eMed" & Tissue == "Biopsy",], trxnCount, by.x ="SeqID",by.y="rn")
rxn_data[,remission_time := paste(Remission, Time_seq, sep ="_")]
rxn_data <- rxn_data[,.(count = sum(value)),by=c("variable","remission_time")]
rxn_data <-  dcast(data = rxn_data, variable~remission_time)
rn <- rxn_data[,variable]
rxn_data <- as.matrix(rxn_data[,-1])
rownames(rxn_data) <- rn
rxn_data <- rxn_data[apply(rxn_data, 1, sum) > 0,]
ca_eMed <- CA(rxn_data, ncp = 20, graph=FALSE) 
fviz_ca_col(ca_eMed)+ ggtitle("eMed")
```

To associate the single reactions with the condition of Remission+Time on can perform a chisquare association for each reaction. Which I will do.

```{r chi2-test}

trxnCount <- data.table(t(rxnCount), keep.rownames=TRUE)
trxnCount <- melt(trxnCount, id.vars = "rn")
rxn_data <- merge(meta[Tissue == "Biopsy",], trxnCount, by.x ="SeqID",by.y="rn")
rxn_data[,remission_time := paste(Remission, Time_seq, sep ="_")]
rxn_data <- rxn_data[,.(count = sum(value)),by=c("variable","remission_time")]
rxn_data <-  dcast(data = rxn_data, variable~remission_time)
rn <- rxn_data[,variable]
rxn_data <- as.matrix(rxn_data[,-1])
rownames(rxn_data) <- rn
rxn_data <- rxn_data[apply(rxn_data, 1, sum) > 0,]

chiMeta <- meta[SeqID %in% colnames(rxnCount),]
smplCounts <-  table(chiMeta[,paste(Remission, Time_seq, sep="_")])

pvals <- c(rep(NA, nrow(rxn_data)))
chi2_mods <- list()
for(i in 1:nrow(rxn_data)){
    rn <- rownames(rxn_data)[i]
    conMat <- matrix(c(rxn_data[i,],smplCounts - rxn_data[i,]), ncol=2)
    chi2 <- chisq.test(conMat)
    chi2_mods[[rn]] <- chi2
    pvals[i] <- chi2[["p.value"]]
}
chi_test <- data.table(
                Rxn = rownames(rxn_data),
                pvalue = pvals,
                padj = p.adjust(pvals, method = "BH"))

# get the residuals for each significant reaction
stdRes <- matrix(0, nrow = nrow(chi_test[padj <= 0.05,]), ncol = ncol(rxn_data),
                 dimnames = list(Rxn = chi_test[padj <= 0.05, Rxn], Condition = colnames(rxn_data)))
for(i in rownames(stdRes)){
    stdRes[i,] <- chi2_mods[[i]]$stdres[,1]
}

rxn_anno <- data.frame(subSystems[,subSys])
rownames(rxn_anno) <- subSystems[,rxns]
colnames(rxn_anno) <- "Subsystem"

pheatmap(stdRes,
         color = colorRampPalette(colors=c("blue","white","red"))(100),
         cluster_cols = FALSE,
         annotation_row = rxn_anno,
         show_rownames = FALSE)

pheatmap(stdRes[,c("C_0","NR_0","R_0")],
         color = colorRampPalette(colors=c("blue","white","red"))(100),
         cluster_cols = FALSE,
         annotation_row = rxn_anno,
         show_rownames = TRUE,
         fontsize = 6)

pheatmap(stdRes[,c("C_0","C_0.167","C_1","C_3","NR_0","NR_0.167","NR_1","NR_3","R_0","R_0.167","R_1","R_3")],
         color = colorRampPalette(colors=c("blue","white","red"))(100),
         cluster_cols = FALSE,
         annotation_row = rxn_anno,
         show_rownames = TRUE,
         fontsize = 6)

DT::datatable(chi_test[padj <= 0.05,])

sigRxn <- rxnCount[chi_test[padj <= 0.05, Rxn],meta[Tissue=="Biopsy",SeqID]]
heatMat <- apply(sigRxn,2,as.integer)
colnames(heatMat) <- colnames(sigRxn)
rownames(heatMat) <- rownames(sigRxn)
col_anno <-as.data.frame(meta[,.(Cohort,
                                 Log10_HB_Mayo = log10(HB_Mayo_impu+1),
                                 Remission,
                                 Log10_Time = log10(Time_seq+1))])
rownames(col_anno) <- meta[,SeqID]

pheatmap(heatMat,
         annotation_col= col_anno,
         #annotation_row = rxn_anno,
         show_colnames = FALSE,
         clustering_distance_rows = "binary",
         clustering_distance_cols = "binary",
         show_rownames = FALSE
)

```

There are `r nrow(chi_test[padj <= 0.05,])` reaction which are associated with an unexpected change of how often this reaction can be found in patients of the respective category. The heatmaps show the standardized residuals, which give a measure how much each condition/reaction contributed to the significance of the chi-square test.

As we are interested in the starting conditions of remitters to non-remitters, I will restrict the following analysis on these conditions only.

```{r chisquare0only }

chiMeta <- meta[SeqID %in% colnames(rxnCount),]
smplCounts <-  table(chiMeta[,paste(Remission, Time_seq, sep="_")])
NR_R_0 <- grep("R_0$",names(smplCounts))

pvals <- c(rep(NA, nrow(rxn_data)))
chi2_mods <- list()
for(i in 1:nrow(rxn_data)){
    rn <- rownames(rxn_data)[i]
    conMat <- matrix(c(rxn_data[i,],smplCounts - rxn_data[i,]), ncol=2)[NR_R_0,]
    chi2 <- fisher.test(conMat)
    chi2_mods[[rn]] <- chi2
    pvals[i] <- chi2[["p.value"]]
}

chi_test_start <- data.table(
                Rxn = rownames(rxn_data),
                pvalue = pvals,
                padj = p.adjust(pvals, method = "BH"))

stdRes <- matrix(0, nrow = nrow(chi_test_start[padj <= 0.05,]), ncol = 2,
                 dimnames = list(Rxn = chi_test_start[padj <= 0.05, Rxn], Condition = colnames(rxn_data)[NR_R_0]))
for(i in rownames(stdRes)){
    stdRes[i,] <- chi2_mods[[i]]$stdres[,1]
}

if(nrow(stdRes) > 2){
    pheatmap(stdRes,
             color = colorRampPalette(colors=c("blue","white","red"))(100),
             cluster_cols = FALSE,
             annotation_row = rxn_anno,
             show_rownames = FALSE)

    DT::datatable(chi_test_start[padj <= 0.05,])
}



```

If I check the starting conditions only and compare remitters to non-remitters, there is are `r nrow(chi_test_start)` significant association in the reactions.

```{r compareResults, fig.width = 7, fig.height = 7}

vennMat <- matrix(0, ncol = 2, nrow = nrow(rxnCount), dimnames = list(Rxn = rownames(rxnCount), Condition = c("Remission_Time","HB_Mayo")))

vennMat[chi_test[padj <= 0.05,Rxn],"Remission_Time"] <- 1
vennMat[wcxTest[padj <= 0.05,rn],"HB_Mayo"] <- 1


vennDiagram(vennMat, circle.col=c("red","blue"))

```

